---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# babelquarto

<!-- badges: start -->
[![R-CMD-check](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of babelquarto is to render a Quarto multilingual book structured like the 
rOpenSci dev guide:

- each qmd is present once for the main language,
- and once more for each other language with an extension à la `.es.qmd`

## Installation

You can install the development version of babelquarto from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ropensci-review-tools/quartobabel")
```

## Example

Create a starter/example book.

```{r}
parent_dir <- withr::local_tempdir()
babelquarto::quarto_multilingual_book(parent_dir = parent_dir, book_dir = "blop")
readLines(file.path(parent_dir, "blop", "_quarto.yml"))
fs::dir_tree(file.path(parent_dir, "blop"))
```

Render it. 
We end up with three books, that cross-link to each other from the left sidebar.

```{r}
babelquarto::render_book(book_path = file.path(parent_dir, "blop"))
fs::dir_tree(file.path(parent_dir, "blop", "_book"))
# if (require("servr") && rlang::is_interactive()) {
#   servr::httw(file.path(parent_dir, "blop", "_book"))
# }
```

Note that this does not _translate_ the content!
Translation tooling will live in [babeldown](https://docs.ropensci.org/babeldown).

### Content translation

From a book whose main language is English...

- Register languages in the Quarto configuration, for instance

```yml
babelquarto:
  mainlanguage: 'en'
  languages: ['es', 'fr']
```

- qmd/Rmd files. `bla.qmd` translation in Spanish would live in `bla.es.qmd`.
- parts. The part title translation can be stored in `_quarto.yml` like so:

```yml
  - part: Building Your Package
    part-es: Construyendo tu paquete
    chapters:
    - pkg_building.Rmd
    - pkg_ci.Rmd
    - pkg_security.Rmd
```

If it does not exist, babelquarto falls back to the part title in the main language.

- title, author, description. Their translation can be stored in `_quarto.yml` like so (NOT in the `book` list):

```yml
book:
  title: Cool book
  author: Myself

title-es: Libro genial
author-es: Yo misma
```

If these fields do not exist, babelquarto falls back to their text in the main language.
