---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# babelquarto

<!-- badges: start -->
[![R-CMD-check](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of babelquarto is to render a Quarto multilingual project, book or website.

The project has to be structured like the 
[rOpenSci dev guide](https://devdevguide.netlify.app/):

- each qmd is present once for the main language,
- and once more for each other language with an extension à la `.es.qmd`; although 
babelquarto will fall back to the file version in the other language.

## Installation

You can install the development version of babelquarto from rOpenSci R-universe:

```r
install.packages('babelquarto', repos = c('https://ropensci.r-universe.dev', 'https://cloud.r-project.org'))
```

Or from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ropensci-review-tools/babelquarto")
```

## Example

Create a starter/example book.

```{r}
parent_dir <- withr::local_tempdir()
book_dir <- "blop"
babelquarto::quarto_multilingual_book(parent_dir = parent_dir, book_dir = book_dir)
readLines(file.path(parent_dir, book_dir, "_quarto.yml"))
fs::dir_tree(file.path(parent_dir, book_dir))
```

```{r, results='hide'}
babelquarto::render_book(file.path(parent_dir, book_dir))
```

We end up with three books, that cross-link to each other from the left sidebar.
[Example](https://devdevguide.netlify.app).

Note that this does not _translate_ the content!
Translation tooling will live in [babeldown](https://docs.ropensci.org/babeldown).

### Configure the base URL

Use the [usual Quarto field](https://quarto.org/docs/websites/website-tools.html), or use the `site_url` argument of `babelquarto::render_book()`.

```
book:
  site-url: https://example.com
```

### Content translation

From a book whose main language is English...

```{r}
parent_dir <- withr::local_tempdir()
book_dir <- "babelbook"
quarto_bin <- quarto::quarto_path()
withr::with_dir(parent_dir, {
  sys::exec_wait(
    quarto_bin,
    args = c("create-project", book_dir, "--type", "book")
  )
})
```

- Register languages in the Quarto configuration, for instance

```{r}
book_path <- file.path(parent_dir, book_dir)
babelquarto::register_main_language(main_language = "en", book_path = book_path)
babelquarto::register_further_languages(further_languages = "es", book_path = book_path)
```


This is how the config file now looks like:

````{r, results='asis', echo=FALSE}
c(
  "```yaml",
  readLines(file.path(book_path, "_quarto.yml")),
  "```"
) |> 
  paste(collapse = "\n") |> 
  cat()
````


- qmd/Rmd files. `bla.qmd` translation in Spanish would live in `bla.es.qmd`. See [babeldown](https://docs.ropensci.org/babeldown) for getting an automated translation.
- parts. The part title translation can be stored in `_quarto.yml` like so:

```yml
  - part: Building Your Package
    part-es: Construyendo tu paquete
    chapters:
    - pkg_building.Rmd
    - pkg_ci.Rmd
    - pkg_security.Rmd
```

If it does not exist, babelquarto falls back to the part title in the main language.

- title, author, description. Their translation can be stored in `_quarto.yml` like so (NOT in the `book` list):

```yml
book:
  title: Cool book
  author: Myself

title-es: Libro genial
author-es: Yo misma
```

If these fields do not exist, babelquarto falls back to their text in the main language.
