---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# babelquarto

<!-- badges: start -->
[![R-CMD-check](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci-review-tools/quartobabel/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of babelquarto is to render a Quarto multilingual project, book or website.

## Installation

You can install the development version of babelquarto from rOpenSci R-universe:

```r
install.packages('babelquarto', repos = c('https://ropensci.r-universe.dev', 'https://cloud.r-project.org'))
```

Or from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ropensci-review-tools/babelquarto")
```

## Example book

Create a starter/example book.

```{r}
parent_dir <- withr::local_tempdir()
project_dir <- "blop"
babelquarto::quarto_multilingual_book(parent_dir = parent_dir, project_dir = project_dir)
readLines(file.path(parent_dir, project_dir, "_quarto.yml"))
fs::dir_tree(file.path(parent_dir, project_dir))
```

```{r, results='hide'}
babelquarto::render_book(file.path(parent_dir, project_dir))
```

We end up with three books, that cross-link to each other from the left sidebar.
[Example](https://devdevguide.netlify.app).


## Example website


Create a starter/example website.

```{r}
parent_dir <- withr::local_tempdir()
project_dir <- "blop"
babelquarto::quarto_multilingual_website(parent_dir = parent_dir, project_dir = project_dir)
readLines(file.path(parent_dir, project_dir, "_quarto.yml"))
fs::dir_tree(file.path(parent_dir, project_dir))
```

```{r, results='hide'}
babelquarto::render_website(file.path(parent_dir, project_dir))
```

[Example](https://maelle.github.io/babelsite), [source](https://github.com/maelle/babelsite)


## Configure the base URL

Use the [usual Quarto field](https://quarto.org/docs/websites/website-tools.html), or use the `site_url` argument of `babelquarto::render_book()`.

```
book:
  site-url: https://example.com
```

## Configure the version text

If you want the choice to be between, say "English" and "Español" rather than "Version in EN" and "Version in ES", add these fields under the babelquarto YAML key, in `_quarto.yml`:

```yaml
babelquarto:
  languagecodes:
  - name: es
    text: "Español"
  - name: en
    text: "English"
```

Using `babelquarto::register_main_language()` and `babelquarto::register_further_languages()` will create the boilertemplate for these fields.

## Configure templates

If you want, say, a banner with different content in the different languages, you can create a [partial template](https://quarto.org/docs/journals/templates.html#html-partials) and in it use the variables `lang-fr` and `lang-en` for instance for content in French and English.

The Quarto configuration file `_quarto.yml` would include:

```yaml
format:
  html:
    template-partials:
      - metadata.html
```

The end of the `metadata.html` template would be:

```html
<!-- copy content from https://github.com/quarto-dev/quarto-cli/blob/main/src/resources/formats/html/pandoc/metadata.html -->
<div class="alert alert-info alert-dismissible">
$if(lang-en)$Hello$endif$
$if(lang-fr)$Salut$endif$
$if(lang-es)$Hola$endif$
</div>
```

## Content translation

From a book whose main language is English...

```{r}
parent_dir <- withr::local_tempdir()
project_dir <- "babelbook"
quarto_bin <- quarto::quarto_path()
withr::with_dir(parent_dir, {
  sys::exec_wait(
    quarto_bin,
    args = c("create-project", project_dir, "--type", "book")
  )
})
```

- Register languages in the Quarto configuration, for instance

```{r}
project_path <- file.path(parent_dir, project_dir)
babelquarto::register_main_language(main_language = "en", project_path = project_path)
babelquarto::register_further_languages(further_languages = "es", project_path = project_path)
```


This is how the config file now looks like:

````{r, results='asis', echo=FALSE}
c(
  "```yaml",
  readLines(file.path(project_path, "_quarto.yml")),
  "```"
) |> 
  paste(collapse = "\n") |> 
  cat()
````


- qmd/Rmd files. `bla.qmd` translation in Spanish would live in `bla.es.qmd`. See [babeldown](https://docs.ropensci.org/babeldown) for getting an automated translation.
- parts. The part title translation can be stored in `_quarto.yml` like so:

```yml
  - part: Building Your Package
    part-es: Construyendo tu paquete
    chapters:
    - pkg_building.Rmd
    - pkg_ci.Rmd
    - pkg_security.Rmd
```

If it does not exist, babelquarto falls back to the part title in the main language.

- title, author, description. Their translation can be stored in `_quarto.yml` like so (NOT in the `book` list):

```yml
book:
  title: Cool book
  author: Myself

title-es: Libro genial
author-es: Yo misma
```

If these fields do not exist, babelquarto falls back to their text in the main language.

Note that babelquarto does not _translate_ the content!
Translation tooling lives in [babeldown](https://docs.ropensci.org/babeldown).
