---
title: "Get started"
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

The goal of this vignette is to show how you can start and maintain a new multilingual Quarto project using {babelquarto}.
There are two types of projects: a book or a website. We will look at each type separately below.

If you want to turn an existing project into a multilingual project, have a look at `vignette("convert")`.

## Installing babelquarto

Before you can start a new multilingual project, you need to install {babelquarto}.

```r
install.packages('babelquarto', repos = c('https://ropensci.r-universe.dev', 'https://cloud.r-project.org'))
```

Or from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("ropensci-review-tools/babelquarto")
```

## Starting a multilingual book

To start a multilingual book, use `quarto_multilingual_book()` with the `parent_dir` argument to specify where you want to create the project and the `project_dir` argument to specify the name of the project.
The argument `main_language` is used to specify the main language of the project and `further_languages` lists all additional languages.

You can set the `site-url` field to your project's URL directly by using the `site_url` parameter of `quarto_multilingual_website()`/`quarto_multilingual_book()`.

```{r}
#| label: setup_book
#| results: 'hide'
parent_dir <- withr::local_tempdir()
project_dir <- "multilingual_book"
babelquarto::quarto_multilingual_book(
  parent_dir = parent_dir,
  project_dir = project_dir,
  main_language = "en",
  further_languages = c("es", "fr")
)
```

Look at the `_quarto.yml` file in the project directory. 
To get familiar with the configuration, take a look at the example below:

```{r}
#| label: display_book_config
#| echo: false
#| results: 'asis'

cat("```{.yaml filename='_quarto.yml'}\n")
readLines(file.path(parent_dir, project_dir, "_quarto.yml")) |>
  cat(sep = "\n")
cat("```\n")
```

The file structure of the project looks like this:

```{r}
#| label: book_dir_structure
fs::dir_tree(file.path(parent_dir, project_dir))
```

Each Quarto file has a Spanish and a French version.
These files aren't automatically translated and are just copies of the original English version.
You will have to provide the translations yourself, or look at {[babeldown](https://docs.ropensci.org/babeldown/)} for automatic translation.
If you look at the `index.qmd` file, you'll see that the French file is called `index.fr.qmd` and the Spanish file is called `index.es.qmd`.

When you're ready to render your book, use `render_book()`.
From within a directory, you don't need to specify the `project_path` argument as it defaults to `.`.

```{r}
#| label: render_book
#| eval: false
withr::with_dir(file.path(parent_dir, project_dir), {
  babelquarto::render_book()
})
```

We end up with three books, that cross-link to each other from the left sidebar.
[Example](https://devguide.ropensci.org/).

By default, in interactive sessions a preview of the project will be opened, 
using the servr package. 
See the [Previewing your multilingual project section](#preview) for further details.

## Starting a multilingual website

To start a multilingual website, use `quarto_multilingual_website()` with the `parent_dir` argument to specify where you want to create the project and the `project_dir` argument to specify the name of the project.
The argument `main_language` is used to specify the main language of the project and `further_languages` lists all additional languages.

You can set the `site-url` field to your project's URL directly by using the `site_url` parameter of `quarto_multilingual_website()`/`quarto_multilingual_book()`.

```{r}
#| label: setup_website
#| results: 'hide'
parent_dir <- withr::local_tempdir()
project_dir <- "multilingual_website"
babelquarto::quarto_multilingual_website(
  parent_dir = parent_dir,
  project_dir = project_dir,
  main_language = "en",
  further_languages = c("es", "fr")
)
```

Look at the `_quarto.yml` file in the project directory. To get familiar with the configuration, take a look at the example below:

```{r}
#| label: display_website_config
#| echo: false
#| results: 'asis'

cat("```{.yaml filename='_quarto.yml'}\n")
readLines(file.path(parent_dir, project_dir, "_quarto.yml")) |>
  cat(sep = "\n")
cat("```\n")
```

Note that the fields like `title-es`, `description-es`, `author-es` are not indented so you could choose to cut and paste them below the default language's `title`, `description`, `author` to more easily remember to update them if needed.

The file structure of the project looks like this:

```{r}
#| label: website_dir_structure
fs::dir_tree(file.path(parent_dir, project_dir))
```

Each Quarto file has a Spanish and a French version.
These files aren't automatically translated and are just copies of the original English version.
You will have to provide the translations yourself, or look at {[babeldown](https://docs.ropensci.org/babeldown/)} for automatic translation.
If you look at the `index.qmd` file, you'll see that the French file is called `index.fr.qmd` and the Spanish file is called `index.es.qmd`.

When you're ready to render your website, use `babelquarto::render_website()`:

```{r}
#| label: render_website
#| eval: false
babelquarto::render_website(file.path(parent_dir, project_dir))
```

We end up with a multilingual website. [Example](https://maelle.github.io/babelsite), [source](https://github.com/maelle/babelsite)

Note that the `_quarto.yml` file created in this process has an example entry set for `site-url:`. Depending on how your multilingual project is being deployed, you may need to either remove or update this entry or alternatively, set the url during rendering by using the `site_url` argument in the render functions (e.g., `babelquarto::render_website(site_url = "https://your.blog.net/your-multilingual-book/")`).

## Previewing your multilingual project {#preview}

Once you have rendered your project, you will have a `_site` or `_book` folder in your project.
In Quarto you would use `quarto preview` to be able to get a look at what you project looks like.
Because of the way {babelquarto} operates, this isn't possible.
You can however preview your files using the [{servr} package](https://cran.rstudio.com/web/packages/servr/index.html).

You can use `servr::httw()` to preview your project.

```{r}
#| eval: false
# For a multilingual website
servr::httw("_site")

# For a multilingual book
servr::httw("_book")
```

This will show an URL that you can open in your IDE or browser to see your project.

If the publishing source for your multilingual project is a 'docs' folder in your repository, you can preview the output with `servr::httw("docs")`.

## Publishing your project

Your Quarto multilingual website or book is a static website.
To publish it, you need to deploy the output folder, i.e. `_site` or `_book`.
That content can be served on any service, for instance GitHub Pages or Netlify.

You cannot rely on usual tooling by Quarto because that would only publish the main language.

You need to publish the entire output folder, including the language folders.
[Example workflow using GitHub Pages by committing the output folder to a gh-pages branch](https://github.com/ropensci/dev_guide/blob/6f9d066d0207e83588a4d9c8dabd93e015083e13/.github/workflows/dev.yml#L58).

## Publishing to GitHub Pages project sites

GitHub Pages supports two common types of sites:

- **User sites**, published at  
  `https://username.github.io/`
- **Project sites**, published at  
  `https://username.github.io/repository-name/`

GitHub Pages **project sites** are a very common way to publish Quarto books and websites. However, they require one extra step when working with multilingual projects.

### Rendering locally vs. final website URLs

When you render a multilingual project **locally** (for example, by running `babelquarto::render_book()` or `babelquarto::render_website()` in RStudio), `{babelquarto}` temporarily removes the site URL. This makes local previewing easier, because the site does not yet live at a real web address.

Because of this, links between languages may be written in a form that works locally but does **not** work once the site is published as a GitHub Pages **project site**.

As a result, a multilingual site can look correct during local preview but have broken language links after deployment.

### How to avoid broken language links on project sites

If you are publishing to a GitHub Pages **project site**, you must make sure that `{babelquarto}` knows the final website address *when the site is rendered*. You can do this in one of two ways.

#### Option 1: Render the site using GitHub Actions (CI)

You can let GitHub build the site automatically. In this case, you provide the final website address using the `BABELQUARTO_CI_URL` environment variable.

This approach is recommended for users who are comfortable setting up GitHub Actions. See `vignette("render-with-ci")` for details.

#### Option 2: Specify the site URL when rendering locally

If you prefer to build the site on your own computer, you can pass the site’s URL directly when rendering, even if it is already listed in `_quarto.yml`:

```r
babelquarto::render_book(
  site_url = "https://username.github.io/repository-name/"
)
```

This ensures that absolute URLs including the repository subpath are generated, and that multilingual navigation works correctly after deployment.

### Summary

- Local renders are designed for previewing and may not include the final website URL.
- GitHub Pages project sites require the final site URL to be known at render time.
- To publish a multilingual project correctly to a project site, you must either:
  - render the site using GitHub Actions, or
  - provide the site URL when rendering locally..
- Without one of these steps, multilingual navigation may break after deployment.

## Next steps

Take a deeper dive into the configuration options available in {babelquarto} and have a look at `vignette("configuration")`.

If you want to translate your multilingual project using automatic translation with DeepL, you should have a look at [babeldown](http://docs.ropensci.org/babeldown/articles/quarto.html).

If you want to setup your own CI and deploy your website, have a look at `vignette("render-with-ci")`.

If you need to personalize the Quarto templates, have a look at `vignette("custom-templates")`.
