---
title: "Multilingual presentations"
vignette: >
  %\VignetteIndexEntry{Multilingual presentations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

The goal of this vignette is to show how you can create and render a multilingual [revealjs](https://quarto.org/docs/presentations/revealjs/) presentation using {babelquarto}.

Multilingual presentations follow the same convention as books and websites (see `vignette("babelquarto")`): each Quarto file exists once per language, with the same basename,

- once with only `.qmd` as extension for the main language (`index.qmd`),
- once with the language code in the extension for each further language (`index.es.qmd`, `index.fr.qmd`).

The difference is that every file declares `format: revealjs`, and that after rendering, each slide deck gets a small fixed button in the top-right corner to switch to the same slide in another language.

## Starting a multilingual presentation

To start a multilingual presentation, use `quarto_multilingual_presentation()` with the `parent_dir` argument to specify where you want to create the project and the `project_dir` argument to specify the name of the project.
The argument `main_language` is used to specify the main language of the project and `further_languages` lists all additional languages.

```{r}
#| label: setup_presentation
#| results: 'hide'
parent_dir <- withr::local_tempdir()
project_dir <- "multilingual_slides"
babelquarto::quarto_multilingual_presentation(
  parent_dir = parent_dir,
  project_dir = project_dir,
  main_language = "en",
  further_languages = c("es", "fr")
)
```

Look at the `_quarto.yml` file in the project directory.
To get familiar with the configuration, take a look at the example below:

```{r}
#| label: display_presentation_config
#| echo: false
#| results: 'asis'

cat("```{.yaml filename='_quarto.yml'}\n")
readLines(file.path(parent_dir, project_dir, "_quarto.yml")) |>
  cat(sep = "\n")
cat("```\n")
```

The file structure of the project looks like this:

```{r}
#| label: presentation_dir_structure
fs::dir_tree(file.path(parent_dir, project_dir))
```

Each Quarto file has a Spanish and a French version.
These files aren't automatically translated and are just copies of the original English version.
You will have to provide the translations yourself, or look at {[babeldown](https://docs.ropensci.org/babeldown/)} for automatic translation.

## Customizing the language button labels

By default the language switch button of each deck is labelled "Version in es", "Version in fr", and so on.
You can change these labels through the `text` field of `languagecodes` in `_quarto.yml`:

```yaml
babelquarto:
  languagecodes:
  - name: es
    text: "Español"
  - name: fr
    text: "Français"
  - name: en
    text: "English"
```

These labels can also be overridden per language through a Quarto profile.
If a `_quarto-es.yml` profile file is present, any `languagecodes` entry it
defines takes precedence over `_quarto.yml` for the Spanish button:

```yaml
babelquarto:
  languagecodes:
  - name: es
    text: "Español (versión completa)"
```

See `vignette("configuration")` for more on profiles.

## Rendering your presentation

When you're ready to render your presentation, use `render_presentation()`:

```{r}
#| label: render_presentation
#| eval: false
babelquarto::render_presentation(file.path(parent_dir, project_dir))
```

We end up with one deck per language in the `_site` folder: the main language at the root, and each further language in its own subfolder (`_site/es`, `_site/fr`).
Every deck contains a button per other language in the top-right corner of the slides.
Clicking a button opens the same slide in the other language.

## Previewing your presentation

Like for books and websites, you can preview the rendered project with the [{servr} package](https://cran.r-project.org/package=servr):

```{r}
#| label: preview_presentation
#| eval: false
servr::httw(file.path(parent_dir, project_dir, "_site"))
```

Note that in interactive sessions, if you do not provide the `site_url` argument, the `render_` functions use relative links, so the language buttons work when the project is served locally by servr.

## Publishing your presentation

A rendered multilingual presentation is a static website: deploy the entire `_site` folder, including the language subfolders, like you would deploy a multilingual website (see `vignette("babelquarto")`).
